A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization

### 2020-10-18 ms/ms.Rmd
pre-release

---
title: "A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization"
runtitle: "Stomatal anatomy and pathogen colonization"
documentclass: frontiersSCNS
author:
  - name: Christopher D. Muir
    affiliation: '1'
    etal: Muir
    email: cdmuir@hawaii.edu
affiliation:
  - id: '1'
    department: School of Life Sciences
    institution: University of Hawaii
    city: Honolulu
    state: Hawaii
    country: USA
output: 
  rticles::frontiers_article:
    includes:
      in_header: header.tex
csl: frontiers.csl
bibliography: stomata-tradeoff.bib
---

@@

<!-- anatomy and gas exchange -->
The stomatal density and maximum pore area set an anatomical upper limit on stomatal conductance [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but stomatal shape, distribution, and patterning also affect gas exchange. Smaller guard cells and dumbbell-shaped stomata of grasses can respond faster to environmental changes [@drake_smaller_2013], but responsiveness is further modulated by subsidiary cell anatomy and physiology [@franks_mechanical_2007; @raissig_mobile_2017; @gray_flanking_2020]. Stomatal clustering reduces gas exchange and photosynthesis because adjacent stomata interfere with one another [@dow_physiological_2014], diffusion shells overlap [@lehmann_effects_2015], and limitations on lateral diffusion of CO$_2$ in the mesophyll [@lawson_stomatal_2014 and references therein]. However, sparse clusters of small stomata could allow a leaf with low rates of gas exchange to have faster stomatal response compared to a leaf with large, low-density stomata [@papanatsiou_stomatal_2017]. Leaves with stomata on both lower and upper surfaces (amphistomatic or amphistomatous) supply more CO$_2$ to the mesophyll than hypostomatic (aka hypostomatous) leaves that only have stomata on the lower surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990; @oguchi_leaf_2018]. In addition to anatomy, the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019]. For example, stomata typically open during the day and close at night in C$_3$/C$_4$ plants, but the opposite is true for CAM plants. Shade, high vapor pressure deficits, dry soil and other factors can cause stomata to (partially) close even in the middle of the day. Variation in how stomata respond to internal and external signals may explain as much of the variation in gas exchange across leaves as anatomy [@lawson_stomatal_2014].

@@

<!-- Two paragraphs focusing the scope of this study -->
Stomatal anatomy could be a key link between gas exchange and pathogen colonization. Although many anatomical factors and stomatal movement affect gas exchange (see above), here I focus on the density and size of stomata in a hypostomatous leaf. Stomatal size refers to both the area of guard cells when fully open, from which one can calculate the pore area for gas exchange (see [Model]). For simplicity, I model a hypostomatous leaf, but consider the implications for amphistomatous leaves in the [Discussion]. Stomatal size and density not only determine the theoretical maximum stomatal conductance (\gsmax), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019]. \gsop{} is the actual stomatal conductance of plants in the field and is almost always below \gsmax{} because stomata are usually not fully open. Although they are not the same, the strong empirical relationship between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used as a proxy for \gsop{} without explicitly modeling dynamic changes in stomatal aperture (see [Discussion]). Stomatal size and density have also been measured on many more species than stomatal responsiveness, which may make it easier to test predictions.

@@

In this section, I introduce a spatially explicit model of successful pathogen colonization on a leaf surface. I explain the model structure and assumptions here; the [Materials and Methods] section below describes how I analyzed the model to address the goals of the study. For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate. The model is agnostic to the type of pathogen (virus, bacterium, fungus, etc.) and the specific biological details of how it moves. For example, motile bacterial cells can land and move around [@beattie_secret_1995] whereas fungi may germinate from a cyst and grow until they form an appresorium for infection [@tucker_surface_2001]. These very different tropic movements on the leaf are treated identically here. I do not model photosynthesis explicitly, but assume that stomatal conductance limits carbon fixation, even though the relationship is nonlinear. I used Sympy version 1.6.1 [@meurer_sympy:_2017] for symbolic derivations.

@@

Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015]. Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$ on the abaxial surface only, since most leaves are hypostomatoues [@muir_making_2015; but see [Discussion]]. This assumption ignores veins, trichomes, and within-leaf variation in stomatal density. Stomata are therefore arrayed in an evenly spaced grid (Figure \ref{fig:fig1}a). The interstomatal distance $U$, measured as the distance from the center of one stomata to the next, is the maximal diagonal of the hexagon in $\mu\textrm{m}$ that forms an equal area boundary between neighboring stomata. The area of a hexagon is $A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2$. By definition the stomatal density is the inverse of this area, such that $D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}$. Therefore, interstomatal distance can be derived from the stomatal density as:

@@

The probability of colonization ($p_\text{colonize}$) is not simply proportional to stomatal cover (\fs). At low \fs, $p_\text{colonize}$ increases rapidly relative to \fs{} at first Figure \ref{fig:fig3}a). At higher \fs, $p_\text{colonize}$ increases linearly with \fs. When pathogens persist indefinitely ($H = 0$), any combination of stomatal size ($S$) and density ($D$) with the same \fs{} have the same effect on $p_\text{colonize}$. When $H > 0$, pathogens are less likely to land close enough to a stomate to infect before dying, so $p_\text{colonize}$ is closer to \fs{} (Figure \ref{fig:fig3}a). The maximum $p_\text{colonize}$ under the range of parameters considered was $\sim 0.6$ when $H = 0$ and \fs{} is at its maximum value of 1/3. When \fs{} is low, $p_\text{colonize}$ is also low. The relationship between $p_\text{colonize}$, \fs, and \gsmax is qualitatively similar in the spatially implicit model, but the values for $p_\text{colonize}$ are substantially higher because pathogens can potentially colonize any stomate on the leaf rather than only those in the focal triangle (see [Appendix 1: Spatially implicit model] for more detail). Bear in mind that this is the probability for a single individual searching randomly; if enough individuals reach the leaf and/or they can actively find stomata, it’s almost certain that at least some will colonize the leaf. However, reducing $p_\text{colonize}$ may help plants limit the damage since fewer total individual pathogens will colonize the leaf interior.

@@

 Model Prediction | How to test it
------------------|----------------
Increasing stomatal size and/or density will have a larger effect on pathogen colonization in leaves with low stomatal cover. | Compare the effect of changing stomatal size and/or density on pathogen colonization in leaves with low and high stomatal cover.
Increasing stomatal size and/or density will have a smaller effect on pathogen colonization in ephiphytic environments more hostile to pathogens. | Compare the effect of changing stomatal size and/or density on pathogen colonization in more hostile environments (e.g. drier, higher light)
When selection against pathogen colonization stronger, the stomatal size-density scaling exponent should be lower | Measure stomatal anatomy in environments that differ in pathogen colonization using comparative or experimental approaches

>>>>

---
title: "A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization"
runtitle: "Stomatal anatomy and pathogen colonization"
documentclass: frontiersSCNS
author:
  - name: Christopher D. Muir
    affiliation: '1'
    etal: Muir
    email: cdmuir@hawaii.edu
affiliation:
  - id: '1'
    department: School of Life Sciences
    institution: University of Hawai'i
    city: Honolulu
    state: Hawaii
    country: USA
output: 
  rticles::frontiers_article:
    includes:
      in_header: header.tex
csl: frontiers.csl
bibliography: stomata-tradeoff.bib
---

@@

<!-- anatomy and gas exchange -->
The stomatal density and maximum pore area set an anatomical upper limit on stomatal conductance [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but stomatal shape, distribution, and patterning also affect gas exchange. Smaller guard cells and dumbbell-shaped stomata of grasses can respond faster to environmental changes [@drake_smaller_2013], but responsiveness is further modulated by subsidiary cell anatomy and physiology [@franks_mechanical_2007; @raissig_mobile_2017; @gray_flanking_2020]. Stomatal clustering reduces gas exchange and photosynthesis because adjacent stomata interfere with one another [@dow_physiological_2014], diffusion shells overlap [@lehmann_effects_2015], and limitations on lateral diffusion of CO$_2$ in the mesophyll [@lawson_stomatal_2014 and references therein]. However, sparse clusters of small stomata could allow a leaf with low rates of gas exchange to have faster stomatal response compared to a leaf with large, low-density stomata [@papanatsiou_stomatal_2017]. Leaves with stomata on both lower and upper surfaces (amphistomatous) supply more CO$_2$ to the mesophyll than hypostomatous leaves that only have stomata on the lower surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990; @oguchi_leaf_2018]. In addition to anatomy, the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019]. For example, stomata typically open during the day and close at night in C$_3$/C$_4$ plants, but the opposite is true for CAM plants. Shade, high vapor pressure deficits, dry soil and other factors can cause stomata to (partially) close even in the middle of the day. Variation in how stomata respond to internal and external signals may explain as much of the variation in gas exchange across leaves as anatomy [@lawson_stomatal_2014].

@@

<!-- Two paragraphs focusing the scope of this study -->
Stomatal anatomy could be a key link between gas exchange and pathogen colonization. Although many anatomical factors and stomatal movement affect gas exchange (see above), here I focus on the density and size of stomata in a hypostomatous leaf. Stomatal size refers to both the area of guard cells when fully open, from which one can calculate the pore area for gas exchange (see [Model]). For simplicity, I model a hypostomatous leaf, but consider the implications for amphistomatous leaves in the [Discussion]. Stomatal size and density not only determine the theoretical maximum stomatal conductance (\gsmax), but are also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019]. \gsop{} is the actual stomatal conductance of plants in the field and is almost always below \gsmax{} because stomata are usually not fully open. Although they are not the same, the strong empirical relationship between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used as a proxy for \gsop{} without explicitly modeling dynamic changes in stomatal aperture (see [Discussion]). Stomatal size and density have also been measured on many more species than stomatal responsiveness, which may make it easier to test predictions.

@@

For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate. The model is agnostic to the type of pathogen (virus, bacterium, fungus, etc.) and the specific biological details of how it moves. For example, motile bacterial cells can land and move around [@beattie_secret_1995] whereas fungi may germinate from a cyst and grow until they form an appresorium for infection [@tucker_surface_2001]. These very different tropic movements on the leaf are treated identically here. I do not model photosynthesis explicitly, but assume that stomatal conductance limits carbon fixation, even though the relationship is nonlinear. I used Sympy version 1.6.1 [@meurer_sympy:_2017] for symbolic derivations.

@@

Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015]. Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$ on the abaxial surface only, since most leaves are hypostomatous [@muir_making_2015; but see [Discussion]]. This assumption ignores veins, trichomes, and within-leaf variation in stomatal density. Stomata are therefore arrayed in an evenly spaced grid (Figure \ref{fig:fig1}a). The interstomatal distance $U$, measured as the distance from the center of one stomata to the next, is the maximal diagonal of the hexagon in $\mu\textrm{m}$ that forms an equal area boundary between neighboring stomata. The area of a hexagon is $A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2$. By definition the stomatal density is the inverse of this area, such that $D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}$. Therefore, interstomatal distance can be derived from the stomatal density as:

@@

The probability of colonization ($p_\text{colonize}$) is not simply proportional to stomatal cover (\fs). At low \fs, $p_\text{colonize}$ increases rapidly relative to \fs{} at first (Figure \ref{fig:fig3}a). At higher \fs, $p_\text{colonize}$ increases linearly with \fs. When pathogens persist indefinitely ($H = 0$), any combination of stomatal size ($S$) and density ($D$) with the same \fs{} have the same effect on $p_\text{colonize}$. When $H > 0$, pathogens are less likely to land close enough to a stomate to infect before dying, so $p_\text{colonize}$ is closer to \fs{} (Figure \ref{fig:fig3}a). The maximum $p_\text{colonize}$ under the range of parameters considered was $\sim 0.6$ when $H = 0$ and \fs{} is at its maximum value of 1/3. When \fs{} is low, $p_\text{colonize}$ is also low. The relationship between $p_\text{colonize}$, \fs, and \gsmax is qualitatively similar in the spatially implicit model, but the values for $p_\text{colonize}$ are substantially higher because pathogens can potentially colonize any stomate on the leaf rather than only those in the focal triangle (see [Appendix 1: Spatially implicit model] for more detail). Bear in mind that this is the probability for a single individual searching randomly; if enough individuals reach the leaf and/or they can actively find stomata, it’s almost certain that at least some will colonize the leaf. However, reducing $p_\text{colonize}$ may help plants limit the damage since fewer total individual pathogens will colonize the leaf interior.

@@

 Model Prediction | How to test it
------------------|----------------
Increasing stomatal size and/or density will have a larger effect on pathogen colonization in leaves with low stomatal cover. | Compare the effect of changing stomatal size and/or density on pathogen colonization in leaves with low and high stomatal cover.
Increasing stomatal size and/or density will have a smaller effect on pathogen colonization in ephiphytic environments more hostile to pathogens. | Compare the effect of changing stomatal size and/or density on pathogen colonization in more hostile environments (e.g. drier, higher light)
When selection against pathogen colonization is stronger, the stomatal size-density scaling exponent should be lower | Measure stomatal anatomy in environments that differ in pathogen colonization using comparative or experimental approaches
